---
title: "BIS 620 Final Project"
author: Anmol Seth, anmol.seth@yale.edu; Julia Stiller, julia.stiller@yale.edu
format:
  html:
    self-contained: true
---
Data Source: ([NCT00364013](https://clinicaltrials.gov/ct2/show/NCT00364013?term=NCT00364013&draw=2&rank=1)). 

```{r Setup, message=FALSE, warning=FALSE}
library(haven)
library(purrr)
library(dplyr)
library(ggplot2)
library(knitr)
library(survival)
library(survminer)

trial_path = file.path("..", "Final Project", "NCT00364013")
trial_files = list.files(trial_path, pattern = "*_pds2019.sas7bdat")

dl = map(file.path(trial_path, trial_files), ~ read_sas(.x))
names(dl) = gsub("*_pds2019.sas7bdat", "", trial_files)

adae = dl$adae #"adverse events" contains severity code 
adsl = dl$adsl #"subject-level" contains ATRT (actual treatment)
adrsp = dl$adrsp #"response" contains confirmed responses
biomark = dl$biomark #contains biomarkers
```

## Visualize Distribution of Genotypes (a.k.a. biomarkers)

```{r}
filtered_biomark <- biomark %>%
  select(c(1:7, 16:17))  # Selecting columns relevant to analysis

# Function to determine genotype
get_genotype <- function(subjid) {
  sublist <- as.list(filtered_biomark[filtered_biomark$SUBJID == subjid, ][, c(3, 5, 7, 9)])
  if ("Mutant" %in% sublist) {
    return("Mutant")
  } else if ((sum(sublist == "Wild-type") > sum(sublist == "Unknown")) | 
             (sum(sublist == "Wild-type") > sum(sublist == "Failure")) |
             (sum(sublist == "Wild-type") > sum(sublist == "")) # OH: need to include empty set
             ) {
    return("Wild-type")
  } else {
    return("Unknown")
  }
}

filtered_biomark$Genotype <- sapply(filtered_biomark$SUBJID, get_genotype)

summary_df <- as.data.frame(table(filtered_biomark$Genotype))
colnames(summary_df) <- c("Category", "Count")

kable(summary_df)
```

```{r}
ggplot(summary_df, aes(x = Category, y = Count, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "KRAS Mutation Status", x = "Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  geom_text(aes(label = Count), vjust = -0.5)
```

should we include predictive/ prognostic analysis done in HW2 to show where we're picking up from?
we know from HW2 that genotype is prognostic but not predictive (genotype impacts death outcome but this is not exacerbated by treatment). so we know that mutants are more likely to die (prognostic), but don't know how many fewer days they will live. And we know that mutants are not more likely to die when being treated (predictive), but we do not know how much longer mutant & wild-types live while on treatment. We set out to answer these questions:


## Visualize how many deaths occured during the study and how long people lived, by genotype

```{r, message=FALSE}
d = left_join(filtered_biomark, 
              adsl[, c("SUBJID", "ATRT", "DTH", "DTHDY")], 
              by = "SUBJID")

d |>
  filter(Genotype != "Unknown") %>% #ignore unknown genotypes
  group_by(Genotype, DTH) |>
  summarize(count = n(), avg_survival_days = sum(DTHDY)/n()) |>
  kable()
```

## Analyze whether number of survival days differed by genotype

```{r}
mutant_survival_days <- d |>
  filter(Genotype == "Mutant") |>
  select(DTHDY)

wild_type_survival_days <- d |>
  filter(Genotype == "Wild-type") |>
  select(DTHDY)

t.test(mutant_survival_days, wild_type_survival_days, var.equal = TRUE)
```

Patients with a mutant genotype generally have 48 to 146 fewer days to live than those with a wild-type (p<0.05).


## Visualize how many deaths occured during the study and how long people lived, by treatment

```{r, message=FALSE}
d |>
  filter(Genotype != "Unknown") %>% #ignore unknown genotypes
  group_by(ATRT, DTH) |>
  summarize(count = n(), avg_survival_days = sum(DTHDY)/n()) |>
  kable()
```


## Analyze whether number of survival days differed by treatment

```{r, message=FALSE}
folfox_survival_days <- d |>
  filter(ATRT == "FOLFOX alone") |>
  select(DTHDY)

folxfox_panitumumab_survival_days <- d |>
  filter(ATRT == "Panitumumab + FOLFOX") |>
  select(DTHDY)

t.test(folfox_survival_days, folxfox_panitumumab_survival_days, var.equal = TRUE)
```

Patients on one or both treatments do not have a different number of survival days (p>0.05).


**Does the combination of Genotype and Treatment Arm have a significant impact on patient survival outcomes?**

```{r}
# Filtering out unknown genotypes since these are not informative for this analysis
d <- subset(d, Genotype != "Unknown")

d$Genotype = factor(d$Genotype, ordered = FALSE)
d$Genotype = relevel(d$Genotype, ref = "Wild-type")

d$ATRT = factor(d$ATRT, ordered = FALSE)
d$ATRT = relevel(d$ATRT, ref = "FOLFOX alone")

surv_plot <- ggsurvplot(
  survfit(Surv(DTHDY, DTH) ~ Genotype + ATRT, data = d),
  data = d,
  conf.int = TRUE,
  legend.title = "Genotype & ATRT",
  title = "Kaplan-Meier Survival Curve by Genotype & Treatment Arm",
  xlab = "Time in Days",
  ylab = "Survival Probability",
  pval = TRUE
)

surv_plot$plot + geom_hline(yintercept = 0.5, col = 'red')
```

Here we can see the following:

- There are some statistically significant interactions with p < 0.0001 although it is not specified what those interactions are.

- Subjects with a "Wild-type" genotype generally show higher survival rates over time. 

- Subjects with a "Mutant" genotype generally show lower survival rates over time.

- If a subject has the "Wild-type" genotype, they may have an improved survival rate if they receive 'Panitumumab+FOLFOX' instead of 'FOLFOX alone'

- If a subject has the "Mutant" genotype, they may have an improved survival rate if they receive 'FOLFOX alone' instead of 'Panitumumab+FOLFOX'

**"Does the presence of KRAS mutations significantly impact the survival outcomes of metastatic colorectal cancer patients receiving different treatment regimens, specifically Panitumumab in combination with FOLFOX and FOLFOX alone?"**

The Kaplan-Meier curve assesses whether there are significant differences in survival times among different genotypic groups .

```{r}
ggsurvplot(
  survfit(Surv(DTHDY, DTH) ~ Genotype, data = d),
  data = d,
  conf.int = TRUE,
  legend.title = "Genotype",
  title = "Kaplan-Meier Survival Curve by Genotype",
  xlab = "Time in Days",
  ylab = "Survival Probability",
  risk.table = TRUE
)
```

```{r}
summary(coxph(Surv(d$DTHDY, d$DTH) ~ d$Genotype))
```

1. **Cox Proportional Hazards Model:**
   - The coefficient for `d$GenotypeMutant` is 0.39534, which is positive. This positive coefficient suggests that having the "Mutant" genotype is associated with a higher hazard (risk) of death compared to the "Wild-type" reference group.

2. **Log-Rank Test:**
   - The log-rank test for the "Genotype" variable has a very small p-value (p=8e-07). This suggests that there is a significant difference in survival between different genotypes.

Patients with the "Mutant" genotype have a lower chance of survival compared to those with the "Wild-type" genotype, as indicated by the positive coefficient and the significant log-rank test.

**Are there differences in the occurrence of adverse events between patients receiving Panitumumab in combination with FOLFOX and those receiving FOLFOX alone?**

```{r}
# Example code
adverse_events <- d %>%
  group_by(ATRT) %>%
  # NOTE: Not sure if this should be 1 or 0
  summarize(AdverseEventRate = mean(DTH == "1", na.rm = TRUE))

ggplot(adverse_events, aes(x = ATRT, y = AdverseEventRate, fill = ATRT)) +
  geom_bar(stat = "identity") +
  labs(title = "Adverse Events by Treatment Arm", x = "Treatment Arm", y = "Adverse Event Rate") +
  theme_minimal()

```

```{r}
# Example code with statistical test
adverse_events <- d %>%
  group_by(ATRT) %>%
  summarize(AdverseEventRate = mean(DTH == "1", na.rm = TRUE))

# Create a contingency table
contingency_table <- table(d$DTH == "1", d$ATRT)

# Perform chi-squared test
chi_squared_test <- chisq.test(contingency_table)

# Print the test results
print(chi_squared_test)

# Plot the bar chart
ggplot(adverse_events, aes(x = ATRT, y = AdverseEventRate, fill = ATRT)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Adverse Events by Treatment Arm",
    x = "Treatment Arm",
    y = "Adverse Event Rate"
  ) +
  theme_minimal()

```
**Are there significant differences in the survival distribution over time between different genotypes?**
This plot should be considered part of the analysis, but merely exploratory


```{r}
ggplot(d, aes(x = DTHDY, color = Genotype)) +
  theme_minimal() +
  geom_density() +
  labs(title = "Survival Distribution Over Time", x = "Time to Death (Days)", y = "Density") +
  theme(legend.position = "top")

```
Are there significant differences in the survival distribution over time between different genotypes?

Hypothesis:
Null Hypothesis (H0): There is no significant difference in the survival distribution over time among different genotypes.
Alternative Hypothesis (H1): There are significant differences in the survival distribution over time among different genotypes.

```{r}
surv_diff_genotype <- survdiff(Surv(DTHDY, DTH) ~ Genotype, data = d)
print(surv_diff_genotype)
```

**Does the treatment arm influence the survival distribution over time within each genotype?**

Hypothesis:
Null Hypothesis (H0): The treatment arm does not significantly influence the survival distribution over time within each genotype.
Alternative Hypothesis (H1): The treatment arm significantly influences the survival distribution over time within each genotype.

```{r}
surv_diff_treatment <- survdiff(Surv(DTHDY, DTH) ~ ATRT, data = d)
print(surv_diff_treatment)
```
